Prediction of Periprosthetic Joint Infection after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden and Denmark

### 2021-02-10 reports/article.Rmd

---
title: "Prediction of Periprosthetic Joint Infection after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden and Denmark"
author: Erik Bulow,^1,2^ Alma Besic Pedersen,^3^ Ina Trolle Andersen,^4^ Ola Rolfson,^1,2^  Nils P. Hailer^5^
date: '`r Sys.Date()`'
bibliography: 
#  H:/zotero_lib.bib
  C:\\Users\\erik_\\Documents\\library.bib
csl: the-lancet.csl
output: 
  bookdown::word_document2:
    reference_docx: article.dotx
    toc: false
    df_print: kable
---

@@

**Correspondance:** [erik.bulow\@vgregion.se](mailto:erik.bulow@vgregion.se){.email}; +46 70 08 234 28; Svenska Hoftprotesregistret, Registercentrum Vastra Gotaland, SE-413 45, Sweden

@@

**OJECTIVE:** Periprosthetic joint infection (PJI) after total hip arthroplasty (THA) is a devastating complication with enormous impact on mortality, quality of life, and healthcare resources. Previous attempts at developing risk calculators are hampered by poor predictive performance of the proposed models or a lack of external validation, and no model is derived on a European population. We thus developed a parsimonious model to predict PJI within 90 days and two years after elective THA.

**PARTICIPANTS AND METHODS:** We analysed `r nmbr(N_shar)` patients recorded in the Swedish Hip Arthroplasty Register `r period` and obtained data on comorbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden. Logistic least absolute shrinkage and selection operator (LASSO) regression was applied to develop prediction models for PJI within 90 days after surgery. Models were internally validated by bootstrapping and externally validated on a cohort of **XXX** patients from **XXX**. We assessed model performance by analysing the area under the receiver operation characteristics curve (AUC) and by calibration belt plots.

@@

The derivation cohort was obtained from patients recorded in the SHAR `r period`. The starting point was chosen because registration of ASA class and body mass index (BMI) in the SHAR was initiated in 2008, and we had no access to data from the national patient register (NPR) after 2015. Data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009; @Cnudde2016] The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI, ASA class, type of hospital, and year of surgery, were obtained from the SHAR. Data on comorbidities for one year prior to index surgery were obtained from the NPR where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to the ICD-10 together with admission and discharge dates. Data on marital status and the educational level of patients was obtained from the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden. The SHAR has completeness of 96-98 % and 100 % coverage,[@Karrholm2018] the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99 % and the positive predictive value is 85-95 %.[@Ludvigsson2011] Death dates were linked from the national population register. Only the last operated hip was considered for patients with bilateral THA.[@Bulow2020]. Patients with missing information on, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a measurement above 50. Patients with missing information on educational levels or the type of hospital performing index surgery were also excluded.

@@

Logistic LASSO regression was performed for each sample using the chosen $\lambda$ as penalty term. Regression coefficients were ranked based on their absolute values as a measure of variable importance. We then used Bayesian methods to estimate a posterior distribution and to identify two change points in a linear segmented piecewise regression model. The second change point was then used as a break-point. Influential predictors with absolute coefficients value above this point, scored a point. The whole process, each with 100 Bootstrap replicates, was repeated 100 times (yielding 10,000 bootstrap samples in total). Potential predictors that were selected at least `r brlasso_selection_n["main"]` out of 100 times were then used in main effects multivariable logistic regression models without penalty and without pre-normalization of numeric variables. Even more parsimonious models with potential predictors chosen at least `r brlasso_selection_n["reduced"]` out of 100 times were evaluated for comparison. Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison. The Rx Risk V index is a comorbidity index based on medical prescriptions coded by the Anatomical Therapeutic Chemical (ATC) classification system during one year prior to surgery. This data was retrieved from the Swedish national prescription register and classified according to Pratt et al.[@Pratt2018] Odds ratios for the final models were estimated with 95 % confidence intervals.

@@

We built an online web calculator available at **XXX** to be used in clinical practice. `r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, decoder, coder and shiny were used.

@@

We included `r nmbr(N_shar)` patients (`r mean(df$P_Sex == "Male") * 100` % males) with a mean age of `r age` years in the derivation cohort from the SHAR (Fig. \@ref(fig:flowchart), left panel). `r crude_rate * 100` % (n = `r format(sum(df$outcome), big.mark = ",")`) of the patients developed a PJI within 90 days (Tab. \@ref(tab:tab1)). The proportion of patients with ASA class III was `r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson comorbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser index, and `r comorb(df$Rx_index_pratt > 0)` had medications classified according to Rx Risk V. In addition, **XXX** patients from **XXX** were included as an external validation cohort (Fig. \@ref(fig:flowchart), right panel).

@@

Since the prediction model was based on easily accessible information it can be translated into a simple formula $\hat p = 1/(1 + e^{-\hat \alpha - \mathbf{\hat \beta X}})$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\mathbf{\beta} = \hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k$ are the estimated intercept and coefficients (Tab. \@ref(tab:coefs)) and $\mathbf{X} = X_1, X_2, \ldots, X_k$ are corresponding patient characteristics. For example, `r examples$desc` for PJI within 90 days.

@@

@@

```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves combines sensitivity and specificity tp illustrate discriminative abilities of the different models. The reduced model perform almost as good as the main model for prediction of PJI within 90 days after surgery, and as good for PJI within 2 years."}
knitr::include_graphics("../graphs/roc.png")
```

@@

```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % bootstrap confidence intervals. The reduced model performs no different than the main model, but both of these perform better than traditional comorbidity measures alone."}

@@

```{r sep, fig.cap = "The vast majority of patients did not have PJI. Hence, the blue bars dominates the histograms (upper panels; note the scales). Normalized density plots reveals however, that patients with PJI had, on average, higher estimated probabilities for this adverse event (lower panels)."}
knitr::include_graphics("../graphs/separation.png")
```

>>>>

---
title: "Prediction of Periprosthetic Joint Infection after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden and Denmark"
author: Erik Bulow,^1,2^ Alma Besic Pedersen,^3^ Ina Trolle Andersen,^4^ Ola Rolfson,^1,2^  Nils P. Hailer^5^ **Ute Hahn + ? + decide order)**
date: '`r Sys.Date()`'
bibliography: 
#  H:/zotero_lib.bib
  C:\\Users\\erik_\\Documents\\library.bib
csl: the-lancet.csl
output: 
  bookdown::word_document2:
    reference_docx: article.dotx
    toc: false
    df_print: kable
    number_sections: false
---

@@

**Correspondance:** [erik.bulow\@gu.se](mailto:erik.bulow@vgregion.se){.email}; +46 70 08 234 28; The Swedish Joint Arthroplasty Register, Registercentrum Vastra Gotaland, SE-413 45, Sweden

@@

**OJECTIVE:** Periprosthetic joint infection (PJI) after total hip arthroplasty (THA) is a devastating complication with enormous impact on mortality, quality of life, and health care resources. Previous attempts at developing risk calculators are hampered by poor predictive performance of the proposed models or a lack of external validation, and no model is derived on a European population. We thus developed a parsimonious model to predict PJI within 90 days after elective THA.

**PARTICIPANTS AND METHODS:** We analyzed `r nmbr(N_shar)` patients recorded in the Swedish Hip Arthroplasty Register `r period` and obtained data on comorbidities and socioeconomic background variables from the National Patient Register and Statistics Sweden. Logistic least absolute shrinkage and selection operator (LASSO) regression was applied to develop prediction models for PJI within 90 days after surgery. Models were internally validated by bootstrapping and externally validated on a cohort of **XXX** patients from The Danish Hip Arthroplasty Register. We assessed model performance by the area under the receiver operation characteristics curve (AUC) and by calibration belt plots.

@@

The derivation cohort was obtained from patients recorded in the SHAR `r period`. The starting point was chosen because registration of ASA class and body mass index (BMI) in the SHAR was initiated in 2008, and we had no access to data from the national patient register (NPR) after 2015. Data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009; @Cnudde2016] The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI, ASA class, type of hospital, and year of surgery, were obtained from the SHAR. Data on comorbidities for one year prior to index surgery were obtained from the NPR where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to the ICD-10 together with admission and discharge dates. Data on marital status and the educational level of patients was obtained from the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden. The SHAR has completeness of 96-98 % and 100 % coverage,[@Karrholm2018] the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99 % and the positive predictive value is 85-95 %.[@Ludvigsson2011] Death dates were linked from the national population register. Only the last operated hip was considered for patients with bilateral THA.[@Bulow2020] Patients with missing information on, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a measurement above 50. Patients with missing information on educational levels or the type of hospital performing index surgery were also excluded.

@@

Logistic LASSO regression was performed for each sample using the chosen $\lambda$ as penalty term. Regression coefficients were ranked based on their absolute values as a measure of variable importance. We then used Bayesian methods to estimate a posterior distribution and to identify two change points in a linear segmented piecewise regression model. The second change point was then used as a break-point. Influential predictors with absolute coefficients value above this point, scored a point. The whole process, each with 100 Bootstrap replicates, was repeated 100 times (yielding 10,000 bootstrap samples in total). Potential predictors that were selected at least `r brlasso_selection_n["main"]` out of 100 times were then used in a main effects multivariable logistic regression model without penalty and without pre-normalization of numeric variables. An even more parsimonious model with potential predictors chosen at least `r brlasso_selection_n["reduced"]` out of 100 times were evaluated for comparison. Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison. The Rx Risk V index is a comorbidity index based on medical prescriptions coded by the Anatomical Therapeutic Chemical (ATC) classification system during one year prior to surgery. This data was retrieved from the Swedish national prescription register and classified according to Pratt et al.[@Pratt2018] Odds ratios for the final models were estimated with 95 % confidence intervals.

@@

We built an online web calculator available at **XXX** to be used in clinical practice. `r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages tidyverse, tidymodels, furrr, pROC, decoder, coder[@Bulow2020b] and shiny were used.

@@

We included `r nmbr(N_shar)` patients (`r mean(df$P_Sex == "Male") * 100` % males) with a mean age of `r age` years in the derivation cohort from the SHAR (Fig. \@ref(fig:flowchart), left panel). `r crude_rate * 100` % (n = `r format(sum(df$outcome), big.mark = ",")`) of the patients developed a PJI within 90 days (Tab. \@ref(tab:tab1)). The proportion of patients with ASA class III was `r comorb(df$P_ASA == "III")`; `r comorb(df$CCI_index_quan_original > 0)` had at least one comorbidity according to the Charlson comorbidity index, `r comorb(df$ECI_index_sum_all > 0)` according to Elixhauser index, and `r comorb(df$Rx_index_pratt > 0)` had medications classified according to Rx Risk V. In addition, **XXX** patients from DHR were included as an external validation cohort (Fig. \@ref(fig:flowchart), right panel).

@@

Since the prediction model was based on easily accessible information it can be translated into a simple formula $\hat p = 1/(1 + e^{-\hat \alpha - \mathbf{\hat \beta X}})$ where $\hat p$ is the estimated probability, $\hat \alpha$ and $\mathbf{\beta} = \hat\beta_1, \hat\beta_2, \ldots, \hat\beta_k$ are the estimated intercept and coefficients (Tab. \@ref(tab:coefs)) and $\mathbf{X} = X_1, X_2, \ldots, X_k$ are corresponding patient characteristics. For example, `r examples$desc` for PJI within 90 days. Note, however, that observed proportions above 10 % were rare. Predicted probabilities above this limit are therefore subject to extrapolation.

@@

Although, not a strict limitation, it should be noted that the aim of prediction is different from traditional statistical inference. As such, confidence intervals presented for the model coefficients (Table \@ref(tab:coefs)) were based solely on the final logistic regression model. Additional uncertainty introduced by the variable selection process was not incorporated. Hence, those intervals should not be interpreted in support of any implicit hypothesis testing concerning the relevance of any individual variables. Also, the individual odds ratios should not be interpreted as effect measures of individual relations. Some variables are correlated. It is therefore not always possible to hold all other variables constant while changing only the relevant one.

@@

```{r rocs, fig.cap = "Receiver Operation Characteristics (ROC) curves combines sensitivity and specificity to illustrate discriminative abilities of the different models. The main and reduced models performed almost identival for prediction of PJI within 90 days after surgery."}
knitr::include_graphics("../graphs/roc.png")
```

@@

```{r aucci, fig.cap = "Area Under the Receiver Operation Characteristics Curve (AUC) as a measure of predictive discriminative ability with 95 % bootstrap confidence intervals. The reduced model performs no different than the main model, and both of these perform better than traditional comorbidity measures alone."}

@@

```{r sep, fig.cap = "The vast majority of patients did not have PJI. Hence, the blue bars dominates the histograms (left panel; note the scales). A normalized density plot reveals however, that patients with PJI had, on average, higher estimated probabilities for this adverse event (lower panels)."}
knitr::include_graphics("../graphs/separation.png")
```
